Shapes analysis for time series.

### 2024-05-15 NEURIPS24/paper.tex
update

Our goal is to analyze the inter-individual variability of a time series dataset, which is of prime interest in medicine and biology \cite{guscelli2019importance,wang2016research,bar2012studying,germain2023unsupervised}.
More specifically, we aim to find an unsupervised features representation method which encode the specificity of an individual compared to another. 
In physiology, studying the different \textit{shapes} in a time series related to biological phenomena and their variations according to individual or pathology is common.
However, a \textit{shape} has no clear definition; it is more an intuitive way to speak about the silhouette of a pattern in a time series. In this paper, we refer to as the shape of a time series, the graph of this signal.

@@

%This type of question has already been addressed in shape analysis \cite{vaillant2004statistics}, a community focusing on the statistical analysis of various mathematical objects invariant under rotations, dilations, or time parameterization.
%However, time series data with their specific structure has not been properly adressed.

 First, we propose not to see time series through their curve $\{s_t:\eqsp t\in\msi\}$, but through their graph $\msg(s)=\{(t,s(t)):\eqsp t\in \msi\} $. 
 Then, building on the shape analysis litterature \cite{beg2005computing,vaillant2004statistics}, we follow the Large Deformation Diffeomorphic Metric Mapping (LDDMM) framework \cite{beg2005computing,vaillant2004statistics} to analyze these graphs.
  The idea is to represent each element $(\msg(s^j))_{j\in[N]}$ of the dataset as the transformation of a reference graph $\msg(\mathbf{s}_0)$ by a diffeomorphism.
  Then, the diffeomorphism is learned by integrating an ordinary differential equation parameterized by a Reproducing Kernel Hilbert Space (RKHS).
   The parameters $(\alpha_j)_{j\in[N]}$ encoding the diffemorphisms $(\phi_j)_{j\in[N]}$ yield the representation features of the graphs $(\msg(s^j))_{j\in[N]}$.
    Finally, these features encoding the shapes can feed any statistical or machine learning model as in URL.

   However, a graph time series transformation by a general diffeomorphism is not always a graph time series, see e.g. \Cref{fig:diffeo}, thus a graph time series is more than a simple curve \cite{glaunes2008large}.
   Our contributions arise from this observation.
   %In this regard, this work is not an application of LDDMM to 1D curves \cite{glaunes2008large}.
    We solve this issue by specifying the class of diffeomorphisms to consider and showing how to learn them.
     This change is fruitful in representing time transformation as illustrated in \Cref{fig:transport}.
     %In particular, the method can handle \textit{irregularly sampled} time series with \textit{variable sizes}.

     Our contributions can be summarized as follows:
     \begin{itemize}
       \item We propose an unsupervised method (TS-LDDMM) to analyze inter-individual variability of shapes in a time series dataset. In particular, the method can handle \textit{irregularly sampled} time series with \textit{variable sizes}.
       \item We motivate our extension of LDDMM to time series by introducing a theoretical framework with a representation theorem for time series graph (\Cref{theorem:representation}) and kernels related to their structure (\Cref{lemma:choice_of_kernel_V}).
       \item We demonstrate the identifiability of the model by estimating the true generating parameter of synthetic data, and we highlight the sensitivity of our method with respect to its hyperparameters, also providing guidelines for tuning.
       We highlight the \textit{interpretability} of TS-LDDMM for studying the inter-individual variability in a clinical dataset.
        We illustrate the quantitative interest of the representation on classification tasks on real shape-based datasets.

     \end{itemize}

@@

\thi{
  Our goal is to analyze the inter-individual variability within a time series dataset, an approach of prime interest in physiological contexts \cite{guscelli2019importance,wang2016research,bar2012studying,germain2023unsupervised}. More specifically, we aim to find an unsupervised features representation method that encodes individual time series specificities compared to a reference one.
}

@@

\thi{
  Our work focuses explicitly on learning shape-based representation of time series. First, we propose not to see the shape of a time series through its curve $\{s_t:\eqsp t\in\msi\}$, but rather through its graph $\msg(s)=\{(t,s(t)):\eqsp t\in \msi\}$. Then, building on the shape analysis literature \cite{beg2005computing,vaillant2004statistics}, we follow the Large Deformation Diffeomorphic Metric Mapping (LDDMM) framework \cite{beg2005computing,vaillant2004statistics} to analyze these graphs. The idea is to represent each element $\msg(s^j)$ of a dataset $(s^j)_{j\in[N]}$ as the transformation of a reference graph $\msg(\mathbf{s}_0)$ by a diffeomorphism $\phi_j$ (ie $\msg(s^j) \sim \phi_j . \msg(\mathbf{s}_0)$ ). The diffeomorphism $\phi_j$ is learned by integrating an ordinary differential equation parameterized by a Reproducing Kernel Hilbert Space (RKHS). The parameters $(\alpha_j)_{j\in[N]}$ encoding the diffemorphisms $(\phi_j)_{j\in[N]}$ yield the representation features of the graphs $(\msg(s^j))_{j\in[N]}$. Finally, these shape-encoding features can feed any statistical or machine-learning model.
  }

@@

\thi{
  Our contributions can be summarized as follows:
    \begin{itemize}
       \item We propose an unsupervised method (TS-LDDMM) to analyze the inter-individual variability of shapes in a time series dataset. In particular, the method can handle multivariate time series \textit{irregularly sampled} and with \textit{variable sizes}.
       \item We motivate our extension of LDDMM to time series by introducing a theoretical framework with a representation theorem for time series graph (\Cref{theorem:representation}) and kernels related to their structure (\Cref{lemma:choice_of_kernel_V}).
       \item We demonstrate the identifiability of the model by estimating the true generating parameter of synthetic data, and we highlight the sensitivity of our method concerning its hyperparameters, also providing guidelines for tuning.
       \item We highlight the \textit{interpretability} of TS-LDDMM for studying the inter-individual variability in a clinical dataset.
        \item We illustrate the quantitative interest of such representation on classification tasks on real shape-based datasets.
 \end{itemize}
}

>>>>

 Our goal is to analyze the inter-individual variability within a time series dataset, an approach of prime interest in physiological contexts \cite{guscelli2019importance,wang2016research,bar2012studying,germain2023unsupervised}.
  More specifically, we aim to find an unsupervised features representation method that encodes individual time series specificities compared to a reference one.
In physiology, studying the different \textit{shapes} in a time series related to biological phenomena and their variations according to individual or pathology is common.
However, a \textit{shape} has no clear definition; it is more an intuitive way to speak about the silhouette of a pattern in a time series. In this paper, we refer to as the shape of a time series, the graph of this signal.

@@

%This type of question has already been addressed in shape analysis \cite{vaillant2004statistics}, a community focusing on the statistical analysis of various mathematical objects invariant under rotations, dilations, or time parameterization.
%However, time series data with their specific structure has not been properly adressed.
Our work focuses explicitly on learning shape-based representation of time series. First, we propose not to see the shape of a time series through its curve $\{s_t:\eqsp t\in\msi\}$, but rather through its graph $\msg(s)=\{(t,s(t)):\eqsp t\in \msi\}$.
   Then, building on the shape analysis literature \cite{beg2005computing,vaillant2004statistics}, we follow the Large Deformation Diffeomorphic Metric Mapping (LDDMM) framework \cite{beg2005computing,vaillant2004statistics} to analyze these graphs. The idea is to represent each element $\msg(s^j)$ of a dataset $(s^j)_{j\in[N]}$ as the transformation of a reference graph $\msg(\mathbf{s}_0)$ by a diffeomorphism $\phi_j$, i.e. $\msg(s^j) \sim \phi_j . \msg(\mathbf{s}_0)$ .
    The diffeomorphism $\phi_j$ is learned by integrating an ordinary differential equation parameterized by a Reproducing Kernel Hilbert Space (RKHS).
     The parameters $(\alpha_j)_{j\in[N]}$ encoding the diffemorphisms $(\phi_j)_{j\in[N]}$ yield the representation features of the graphs $(\msg(s^j))_{j\in[N]}$. Finally, these shape-encoding features can feed any statistical or machine-learning model.
%  First, we propose not to see time series through their curve $\{s_t:\eqsp t\in\msi\}$, but through their graph $\msg(s)=\{(t,s(t)):\eqsp t\in \msi\} $. 
%  Then, building on the shape analysis litterature \cite{beg2005computing,vaillant2004statistics}, we follow the Large Deformation Diffeomorphic Metric Mapping (LDDMM) framework \cite{beg2005computing,vaillant2004statistics} to analyze these graphs.
%   The idea is to represent each element $(\msg(s^j))_{j\in[N]}$ of the dataset as the transformation of a reference graph $\msg(\mathbf{s}_0)$ by a diffeomorphism.
%   Then, the diffeomorphism is learned by integrating an ordinary differential equation parameterized by a Reproducing Kernel Hilbert Space (RKHS).
%    The parameters $(\alpha_j)_{j\in[N]}$ encoding the diffemorphisms $(\phi_j)_{j\in[N]}$ yield the representation features of the graphs $(\msg(s^j))_{j\in[N]}$.
%     Finally, these features encoding the shapes can feed any statistical or machine learning model as in URL.

However, a graph time series transformation by a general diffeomorphism is not always a graph time series, see e.g. \Cref{fig:diffeo}, thus a graph time series is more than a simple curve \cite{glaunes2008large}.
 Our contributions arise from this observation: we specify the class of diffeomorphisms to consider and show how to learn them.
  This change is fruitful in representing transformations of time series graphs as illustrated in \Cref{fig:transport}.
   %In this regard, this work is not an application of LDDMM to 1D curves \cite{glaunes2008large}.

     %In particular, the method can handle \textit{irregularly sampled} time series with \textit{variable sizes}.

    %  Our contributions can be summarized as follows:
    %  \begin{itemize}
    %    \item We propose an unsupervised method (TS-LDDMM) to analyze inter-individual variability of shapes in a time series dataset. In particular, the method can handle \textit{irregularly sampled} time series with \textit{variable sizes}.
    %    \item We motivate our extension of LDDMM to time series by introducing a theoretical framework with a representation theorem for time series graph (\Cref{theorem:representation}) and kernels related to their structure (\Cref{lemma:choice_of_kernel_V}).
    %    \item We demonstrate the identifiability of the model by estimating the true generating parameter of synthetic data, and we highlight the sensitivity of our method with respect to its hyperparameters, also providing guidelines for tuning.
    %    We highlight the \textit{interpretability} of TS-LDDMM for studying the inter-individual variability in a clinical dataset.
    %     We illustrate the quantitative interest of the representation on classification tasks on real shape-based datasets.

    %  \end{itemize}
    Our contributions can be summarized as follows:
    \begin{itemize}
       \item We propose an unsupervised method (TS-LDDMM) to analyze the inter-individual variability of shapes in a time series dataset. In particular, the method can handle multivariate time series \textit{irregularly sampled} and with \textit{variable sizes}.
       \item We motivate our extension of LDDMM to time series by introducing a theoretical framework with a representation theorem for time series graph (\Cref{theorem:representation}) and kernels related to their structure (\Cref{lemma:choice_of_kernel_V}).
       \item We demonstrate the identifiability of the model by estimating the true generating parameter of synthetic data, and we highlight the sensitivity of our method concerning its hyperparameters, also providing guidelines for tuning.
       \item We highlight the \textit{interpretability} of TS-LDDMM for studying the inter-individual variability in a clinical dataset.
        \item We illustrate the quantitative interest of such representation on classification tasks on real shape-based datasets.
 \end{itemize}

@@

% \thi{
%   Our goal is to analyze the inter-individual variability within a time series dataset, an approach of prime interest in physiological contexts \cite{guscelli2019importance,wang2016research,bar2012studying,germain2023unsupervised}. More specifically, we aim to find an unsupervised features representation method that encodes individual time series specificities compared to a reference one.
% }

@@

% \thi{
%   Our work focuses explicitly on learning shape-based representation of time series. First, we propose not to see the shape of a time series through its curve $\{s_t:\eqsp t\in\msi\}$, but rather through its graph $\msg(s)=\{(t,s(t)):\eqsp t\in \msi\}$.
%    Then, building on the shape analysis literature \cite{beg2005computing,vaillant2004statistics}, we follow the Large Deformation Diffeomorphic Metric Mapping (LDDMM) framework \cite{beg2005computing,vaillant2004statistics} to analyze these graphs. The idea is to represent each element $\msg(s^j)$ of a dataset $(s^j)_{j\in[N]}$ as the transformation of a reference graph $\msg(\mathbf{s}_0)$ by a diffeomorphism $\phi_j$ (ie $\msg(s^j) \sim \phi_j . \msg(\mathbf{s}_0)$ ).
%     The diffeomorphism $\phi_j$ is learned by integrating an ordinary differential equation parameterized by a Reproducing Kernel Hilbert Space (RKHS).
%      The parameters $(\alpha_j)_{j\in[N]}$ encoding the diffemorphisms $(\phi_j)_{j\in[N]}$ yield the representation features of the graphs $(\msg(s^j))_{j\in[N]}$. Finally, these shape-encoding features can feed any statistical or machine-learning model.
%   }

@@

% \thi{
%   Our contributions can be summarized as follows:
%     \begin{itemize}
%        \item We propose an unsupervised method (TS-LDDMM) to analyze the inter-individual variability of shapes in a time series dataset. In particular, the method can handle multivariate time series \textit{irregularly sampled} and with \textit{variable sizes}.
%        \item We motivate our extension of LDDMM to time series by introducing a theoretical framework with a representation theorem for time series graph (\Cref{theorem:representation}) and kernels related to their structure (\Cref{lemma:choice_of_kernel_V}).
%        \item We demonstrate the identifiability of the model by estimating the true generating parameter of synthetic data, and we highlight the sensitivity of our method concerning its hyperparameters, also providing guidelines for tuning.
%        \item We highlight the \textit{interpretability} of TS-LDDMM for studying the inter-individual variability in a clinical dataset.
%         \item We illustrate the quantitative interest of such representation on classification tasks on real shape-based datasets.
%  \end{itemize}
% }
